\documentclass{article}

\usepackage{amsmath}
\usepackage[a5paper]{geometry}

\title{An Introduction to Object-Oriented Design}
\author{Lane W. Surface}
\date{April 2019}

% \geometry{a5paper}

\begin{document}

\pagenumbering{gobble}
\begin{center}
\vspace*{1.5cm}
\LARGE
\textbf{An Introduction to Object-Oriented Design}

\vspace{0.75cm}
\large
Lane W. Surface 

\vspace{0.35cm}
April 2019
\end{center}

\newpage
\pagenumbering{arabic}
\noindent
\textsf{
Among many of the frameworks, APIs, languages, and other skill sets you will be
expected to know as a professional programmer, a sound knowledge of the
principles of object-oriented design (henceforth referred to as OOD) will be
invaluable in your career. Though it may seem simple on the surface, even
professionals struggle with the decisions that come with a well-designed and
easy to maintain system which strives to follow the best OOD practices. This
document will serve primarily as a brief introduction for developers with no or
limited experience in this domain.
}

\section{Classes, Methods, and State: It All Ties Into Encapsulation}
A recurring theme which you will come to see in the evolution of programming
languages, and thus software development in general, is the tendency for a
language to make the solution to a given problem both easier to write and easier to read.
Assembly languages were originally the de-facto standard for writing
applications, and many were written solely in it, including important and
critical software like that which ran on the Apollo Guidance Computers and put
men on the moon. From these struggles, a new way of writing software was
proposed, and higher-level languages based on the notion of functions were
created. From these languages has come one of the newest developments in the
programming language world, which is based around the notion of the
\textit{object}.
\paragraph{}
But what is an object anyway? For how often the term is thrown around, there
really isn't must of an attempt to explain it or the principles on which it is
based. An object is simply a component of an object-oriented system; this means
that there are many objects which make up an object-based software system. Each
of these objects interacts with the others which make up this application to
create a working program.
\paragraph{}
Let’s expound this definition further: Each object itself has both state and
behavior. This state, often called fields of the object, resides in class-level
variables which are \textit{instance-dependent}. In some languages these are
called
instance variables. State is a property of the object itself. Say you have a
human object. This object might have state such as hair color, race, ethnicity,
nationality, etc.
\paragraph{}
Although the state of an object is important, it is mostly internal (and should
be kept hidden; we'll address this further in a moment). \textit{The primary way that
objects should expose their functionality to other objects in a system is though
instance-dependent functions, which are often referred to as methods of that
object.} These methods use the state of their object to determine the action
that
should be taken, without requiring the objects which call that function to know
about the state itself. Therefore, calling a method such as \texttt{pray()} on a human
object should determine the way that it prays by these factors, without a caller 
needing to know all of the internal details.
\paragraph{}
Now I've skipped over an important distinction to make, as I wanted to avoid
confusing you up front. How do we create and address the objects in our system?
This state is not intrinsic, and we have to define it somewhere in our
application. We do so by defining classes. Classes are a template for the
construction of an object. In other words, the class defines the \textit{how},
and when
we construct an object from a class, we define the \textit{what}. A class will
determine
the kinds of things which we need to know about. Taking from the earlier example
of a human object, the human class will tell the compiler that we need each of
our human objects to have an integer for their age, string for their name, etc.
Then when we initialize a human object (as there may be many in our system), we
provide the values of these fields. It is only after an object has been
constructed and initialized that we can call the methods which it defines. The
type of the object is the same as the class which defines it, and the name that
the variable is given when constructed is the object itself.
\paragraph{}
You may be wondering why it’s considered good practice to not expose the state
of an object. The reasons are twofold. First, forcing a client to interact with
the state of an object increases the complexity of your code. The purpose of
designing an object-oriented system is to hide away information that is not
important to the rest of the system. If the only thing we need to know is what
behavior an object supports (i.e., the methods which it defines in its public
interface), then we are much less likely to be surprised by the code that we
write. Another reason is because, as software is maintained over its lifetime,
the fields and their respective types are very likely to change; it is not often
that we write code once and are done with it. \textit{The majority of the
lifetime of a
program is in maintaining it}. We will see that it is much easier to make
modifications to the state and behavior of an object when we do not expose this
state directly.

\section{Further Conveniences of Object-Oriented Systems}
Although encapsulation leads to reduced system complexity in general, the real
power is not in this principle at all. In this section, we will explore ways to
create cohesive and easily intelligible software in an object-oriented fashion,
and see the way that the design decisions made in the implementation of one of
these systems increases the legibility of complex software overall. Proper and
thoughtful design is difficult; however, with practice, you will see how
object-oriented programming languages can increase developer productivity
significantly when in the right hands.
\subsection{Class-Based Inheritance}
Of all the features available to you in an object-oriented toolset, inheritance
is one of the best-known and most frequently abused aspects of any
object-oriented programming language. Although powerful in its ability to reduce
program size and shared object features, inheritance comes with a price and
should be used carefully.
\paragraph{}
Inheritance is the ability for related classes to share the parts which are
similar between them. If two classes have parts which are identical, they can
each inherit from a common ancestor, which is usually referred to as the
\textit{superclass} of each of them. Each of these objects, in turn, can serve
as the
superclass of another class of objects, and the inheritance hierarchy continues
naturally, as you would expect for it to. For example, the human class which we
defined earlier could decide to inherit from a class named mammal. If we decided
to add another class named cat, this class could also decide to inherit
functionality from the common mammal class. These so-called \textit{subclasses}
can add
their own functionality, yet retain the characteristics which they have in
common. As the complexity of our system grows and more classes are added, the
power of this ability becomes more prevalent as the amount of code is decreased
significantly. That is to say, as the number of subclasses grows, the simpler
it is to understand the system.
\paragraph{}
As mentioned before, there are some drawbacks to this ability as the hierarchy
grows more complex (as there are to most things in object-oriented
programming).If the division between these classes is not chosen carefully, the
class
hierarchy can quickly grow out of control. When this happens, other programmers
who add to this system later are likely to reimplement something that already
exists in one of these superclasses. This drawback mitigates the decision to
employ inheritance in the first place, and can actually make the code base much
more complex than was originally called for.

\subsection{Interfaces and Implementations}
I have mentioned numerous times within this document the “interface” of an
object. What does this mean exactly? An interface is the way that other objects
can interact with another. \textit{Essentially, it is the behavior, or methods,
which an
object makes publicly available for use by any other}. In the event that
encapsulation is used properly, the idea is that knowing an object’s interface,
or the behavior it exposes, is enough to properly use the object itself.
\paragraph{}
This has some advantages in terms of creating robust systems. (When I state that
a system is robust, what I really mean is it is robust to change. This means
that we can add functionality, most likely in the form of new classes, later
down the road without crippling the architecture we have so carefully designed.)
Recalling the information concerning inheritance which was detailed in the
previous section, we know that classes may have inheritance-based relationships
with one another. This is significant in terms of fortifying our system against
change and easing the duty of expanding it in the future.
\paragraph{}
We can provide for objects which share some common behavior a common superclass 
which defines the methods for this behavior. This implies, of course, that any 
objects which are related to that superclass both support the behavior which it 
defines and in the way that it defines it. \textit{We can interact with any object 
that  inherits from this superclass as if it were the superclass itself, and we 
don’t even need to know the class of the object we are dealing with.} If we know 
the interface for a class of objects, we know how to work with them. This 
principle of knowing how to interact with an object without knowing its concrete 
class is known as \textit{transparency}---the idea that our system only sees the 
parts which are important and has the liberty to ignore the rest.
\paragraph{}
Using interfaces in a system to define the behavior which classes support is a
good practice in object-oriented programming. This design practice is called
\textit{programming to an interface} and creates classes which are
decoupled from each other. The state of one class being decoupled from another
means that one can vary without affecting other objects which use it. 
\subsection{Polymorphism; Or How To Predict The Future}
When our class inheritance hierarchy is based around interfaces which define the 
behavior which objects support, we have several other advantages which we need 
to keep in mind. One of these advantages, as we have already noted, is the
decoupling of behavior from the implementation by classes which inherit, or
implement, that interface. When we define the state of our objects, or the
parameters to our methods then, we can use this interface as the object type
instead of the concrete type itself. By doing so, we future-proof ourselves
against changes in the system. 
\paragraph{}
When we design our interfaces well, and classes which support a behavior which
could be subject to change implement an interface, it is a painless process to 
expand this system. Adding new functionality is as simple as writing a new
class which inherits from that same interface and exchanging the instances of
that class whereever it is deemed convenient. The actual implementation in 
this new class may do different things than the original, but it supports
that behavior through the same mechanisms as the old instances.

\section{Object Relationships}
We have seen that classes of objects may declare inheritance-based relationships 
to share code which they have in common, as well as be referenced by this common
type. These inheritance (subtype) relationships are not the only ones we have
available to use. Objects which are related by a common supertype are said to
be related through an \textit{is-a} relationship; this means that they can both
be considered to be of the ancestor class which they share. A car is a type of
vehicle---therefore, in an object-based system, we might write a vehicle 
interface which defines the behavior all should support. Imagine our interface
has methods such as \texttt{move(int amount)} and \texttt{applyBrake()}. We would
expect for both a bicycle and a car to be able to implement these methods, even
if the method which they use to do it is not the same.
\paragraph{}
What about when objects do not have an is-a relationship? What other cases can
objects be related by? Objects which have functionality but do not necessarily
share a superclass with any other (that is to say that some objects may have
functionality that is internal to their operation, but do not expose this 
functionality as a feature of interacting with the object itself). In this case,
an object may have an instance of that functionality, which is implemented by
another class, as state of the object. Methods which are called on this object
may use the functionality of this other object by \textit{delegating} operations 
to its methods; we may call methods of another object as a side-effect of behavior
of the receiver of that invocation. This is known as a \textit{has-a} relationship
(because the object which delegates stores the other object in its state).

\end{document}